The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-27 manuscript/manuscript.Rnw
prep for genome biology submission

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
%\usepackage[plain]{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
\usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\setcounter{tocdepth}{3}
\oddsidemargin 0.0in 
\textwidth 6in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% set a large emergency stretch to avoid issues with overfull boxes
\setlength{\emergencystretch}{20em}
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

@@

\begin{document}
\author[1]{Don Armstrong}
\author[2]{Michael R. McGowen}
\author[1]{Amy Weckle}
\author[3]{Jason Caravas}
\author[4]{Dalen Agnew}
\author[5]{Kurt Benirschke}
\author[6]{Sue Savage-Rumbaugh}
\author[7,8]{Chong Jai Kim}
\author[9]{Günter P. Wagner}
\author[7]{Roberto Romero}
\author[1,6,10,*]{Derek E. Wildman}
\title{The evolution of gene expression in the term placenta of viviparous mammals}
% These are all wrong, but we'll change them later.
\affil[1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\affil[2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\affil[3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\affil[4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\affil[5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\affil[6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\affil[7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\affil[8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\affil[9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\affil[10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\affil[*]{To whom correspondence should be addressed}
\date{\today}

% Acronyms
\input{acronyms}
% \twocolumn
\maketitle
% \onehalfspacing
\doublespacing

@@

\section{Abstract}

The placenta is arguably the most anatomically variable organ in
mammals. Understanding gene expression patterns among placentas from
multiple species can illuminate the molecular basis of placenta
function as well as potentially enrich our knowledge of the biology of
obstetrical syndromes. We profiled the term placenta by using Illumina
RNA-Seq technology to sequence the transcriptomes of
\Sexpr{star.logs[,length(unique(species))]} species of therian (i.e.
marsupials and “placental”) mammals.

@@

\section{Introduction}

@@

\begin{table}
  \begin{tabular}{c c c p{3in}}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

Fetal tissue was collected from the placenta of six mammalian species
(\human [human], \bonobo [bonobo], \dog [domestic dog], \cow [domestic
cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \gallili [blind mole rat from Upper Galilee],
\carmelli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure\cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top
5\% of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology
using the 1:1 \human ortholog obtained

@@

All 1:1 orthologues present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via
Biomart\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the
biomaRt R package \DLA{Cite the biomaRt package}. A total of
\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
analyses of gene expression evolution described below.

@@

\begin{table}
  \tiny
<<primate.lineage.specific,echo=FALSE,results="asis">>=
pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
setkey(pri.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed significantly in the primate
  order as compared to all other analyzed species.}
\label{tab:primate_lineage_specific}
\end{table}

@@

\begin{table}
  \tiny
<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="homo sapiens")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.human.genes,"Gene symbol")
@ 
\caption{The top 50 genes by \ac{FPKM} in \human which are not
  annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
  name corresponding to the human symbol. A complete list of all genes
  is in \cref{tab:all_homo_sapiens_placenta}.}
\label{tab:top50_human_placenta}
\end{table}

@@

\begin{table}
  \tiny
<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
print(xtable(top.bonobo.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.bonobo.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
\label{tab:top50_bonobo_placenta}
\end{table}

@@

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="SERPINE2" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
@ 
\caption{\genetreexplanation{\textit{SERPINE2}}}
\label{fig:serpine_evolution_tree}
\end{figure}

@@

\begin{table}
  \tiny
<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
print(xtable(top.spider.monkey.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.spider.monkey.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
\label{tab:top50_spider_monkey_placenta}
\end{table}

@@

\begin{table}
  \tiny
<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
top.dog.genes <- top.genes.table(species.name="canis familiaris")
print(xtable(top.dog.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.dog.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
\label{tab:top50_dog_placenta}
\end{table}

@@

>>>>

%\documentclass[english,12pt]{article}
\documentclass[doublespacing]{bmcart_dla}
\usepackage{fontspec}
\setmainfont{FreeSans}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage[letterpaper]{geometry}
\usepackage{url}
% \usepackage{fancyhdr}
\usepackage{graphicx}
% \usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
%\usepackage[plain]{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
% \usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\newpage \section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
\usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
% \newenvironment{narrow}[2]{%
%   \begin{list}{}{%
%       \setlength{\topsep}{0pt}%
%       \setlength{\leftmargin}{#1}%
%       \setlength{\rightmargin}{#2}%
%       \setlength{\listparindent}{\parindent}%
%       \setlength{\itemindent}{\parindent}%
%       \setlength{\parsep}{\parskip}}%
%   \item[]}{\end{list}}
% \def\newblock{\hskip}
% \newenvironment{paperquote}{%
%   \begin{quote}%
%      \it
%   }%
%   {\end{quote}}
% \renewcommand{\textfraction}{0.15}
% \renewcommand{\topfraction}{0.85}
% \renewcommand{\bottomfraction}{0.65}
% \renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
% \newenvironment{enumerate*}%
%   {\begin{enumerate}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{enumerate}}
% \newenvironment{itemize*}%
%   {\begin{itemize}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{itemize}}
% \setcounter{tocdepth}{3}
% \oddsidemargin 0.0in 
% \textwidth 6in
% \raggedbottom
% \clubpenalty = 10000
% \widowpenalty = 10000
% set a large emergency stretch to avoid issues with overfull boxes
% \setlength{\emergencystretch}{20em}
% make the afill smaller
% \renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
% \pagestyle{fancy}

@@

\begin{document}
\raggedright

@@

% Acronyms
\input{acronyms}

\begin{frontmatter}
\begin{fmbox}
\dochead{Research}
\title{The evolution of gene expression in the term placenta of viviparous mammals}%
%
\author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
\author[addressref={aff2}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL} \fnm{Amy} \snm{Weckle}}
\author[addressref={aff3}]{\inits{J}\fnm{Jason} \snm{Caravas}}
\author[addressref={aff4}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
\author[addressref={aff5}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
\author[addressref={aff6}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
\author[addressref={aff7,aff8}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
\author[addressref={aff9}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
\author[addressref={aff7}]{\inits{R}\fnm{Roberto} \snm{Romero}}
\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
% These are all wrong, but we'll change them later.
\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
%\address[id=affcor]{Derek Wildman <wildmand@illinois.edu>}

\begin{artnotes}
\end{artnotes}

\end{fmbox}
% \twocolumn

\begin{abstractbox}
\begin{abstract}
% \section{Abstract}
  \parttitle{Background} The placenta is arguably the most
  anatomically variable organ in mammals. Understanding gene
  expression patterns among placentas from multiple species can
  illuminate the molecular basis of placenta function as well as
  potentially enrich our knowledge of the biology of obstetrical
  syndromes.

  \parttitle{Results} We profiled the term placenta by using Illumina
  RNA-Seq technology to sequence the transcriptomes of
  \Sexpr{star.logs[,length(unique(species))]} species of therian (i.e.
  marsupials and “placental”) mammals.

  \parttitle{Conclusions}

@@

\end{abstract}
\end{abstractbox}
\end{frontmatter}

\section{Background}

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

Fetal tissue was collected from the placenta of eight mammalian
species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top
5\% of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology
using the 1:1 \human ortholog obtained

@@

All 1:1 orthologues present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart
\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
package
\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
analyses of gene expression evolution described below.

@@

\begin{table}
  \footnotesize
<<primate.lineage.specific,echo=FALSE,results="asis">>=
pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
setkey(pri.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed significantly in the primate
  order as compared to all other analyzed species.}
\label{tab:primate_lineage_specific}
\end{table}

@@

\begin{table}
  \footnotesize
<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="homo sapiens")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.human.genes,"Gene symbol")
@ 
\caption{The top 50 genes by \ac{FPKM} in \human which are not
  annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
  name corresponding to the human symbol. A complete list of all genes
  is in \cref{tab:all_homo_sapiens_placenta}.}
\label{tab:top50_human_placenta}
\end{table}

@@

\begin{table}
  \footnotesize
<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
print(xtable(top.bonobo.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.bonobo.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
\label{tab:top50_bonobo_placenta}
\end{table}

@@

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="SERPINE2" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
@ 
\caption{\genetreeexplanation{\textit{SERPINE2}}}
\label{fig:serpine_evolution_tree}
\end{figure}

@@

\begin{table}
  \footnotesize
<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
print(xtable(top.spider.monkey.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.spider.monkey.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
\label{tab:top50_spider_monkey_placenta}
\end{table}

@@

\begin{table}
  \footnotesize
<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
top.dog.genes <- top.genes.table(species.name="canis familiaris")
print(xtable(top.dog.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.dog.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
\label{tab:top50_dog_placenta}
\end{table}

@@

\begin{backmatter}
\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
Don Armstrong analyzed all of the data and wrote the paper.

Michael McGowan wrote the paper and performed initial analyses.

\section*{Acknowledgements}
  Text for this section \ldots

\printaddresses

\end{backmatter}
